---
title: "Add Plant Functional Type labels"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## USDA Plants Database

USDA Plants Database R API tutorial: https://recology.info/2016/10/usda-plants-database-r/ 

The USDA maintains a database of plant information, some of it trait data, some of it life history. Check it out at http://plants.usda.gov/java/

```{r install_load_packages, message=FALSE}
#install.packages("request")
#library(devtools)
#devtools::install_github("sckott/request")
library(request)
library(tibble)
library(tidyverse)
library(here)
```

```{r}
# Set the base url
root <- api("https://plantsdb.xyz")

root %>% api_path(heartbeat)
```

^^ My results here do not match those that show up on the tutorial page above. Maybe the API is down or the URL has changed? 

Based on the GitHub issue discussion (https://github.com/USDA/USDA-APIs/issues/7), someone uploaded USDA plants data here from 2016: https://uofi.app.box.com/v/usdaplants

```{r, eval=FALSE, include=FALSE}
# Specify URL where file is stored 
url <- "https://uofi.app.box.com/v/usdaplants/folder/6688524249/usda_plants_20160713.csv"
destFile <- file.path(here::here(), "analysis","data", "usda_plants_20160713.csv")
# Apply download.file function in R
download.file(url, destFile)
```

```{r}
#usda_plants <- read_csv(file.path(here::here(), "analysis", "data", "usda_plants_20160713.csv"),
                        local = locale(encoding = "latin1"))

# Error in nchar(x) : invalid multibyte string, element 2
# ^ Issue with some of the column names in the .csv? 
```



## BIEN 

RBIEN tutorial: https://cmerow.github.io/RDataScience/3_3_RBIEN_tutorial.html

```{r install_load_packages_bien, message=FALSE}
library(BIEN)
library(ape) # Package for working with phylogenies in R
library(maps) # Useful for making quick maps of occurrences
library(sp) # A package for spatial data
```

"Let’s say we’re interested in the species Xanthium strumarium and we’d like some occurrence data."

```{r}
Xanthium_strumarium <- BIEN_occurrence_species(species = "Xanthium strumarium")
#View(Xanthium_strumarium)
head(Xanthium_strumarium)
str(Xanthium_strumarium)
```


BIEN traits and taxonomic data

"Let’s say we’re interested in the genus [Abies], and we’d like to get an idea of how many species there are in this genus and what higher taxa it falls within."

*Abies* genus is fir trees (coniferous trees). The "higher_plant_group" returned by the BIEN taxonomy function is "gymnosperms (conifers)"

```{r}
Abies_taxonomy<-BIEN_taxonomy_genus(genus = "Abies")

#View(Abies_taxonomy)
Abies_taxonomy[1:10,]
```

*Acer* genus is maple trees (deciduous trees). The "higher_plant_group" returned by the BIEN taxonomy is "flowering plants".

```{r}
Acer_taxonomy<-BIEN_taxonomy_genus(genus = "Acer")

#View(Acer_taxonomy)
Acer_taxonomy[1:10,]
```
*Acer rubrum* species is the red maple (deciduous trees). The "higher_plant_group" returned by the BIEN taxonomy is "flowering plants".

```{r}
Acer_rubrum_taxonomy <- BIEN_taxonomy_species(species = "Acer rubrum")
Acer_rubrum_taxonomy[1:10,]
```
```{r}
Artemisia_tridentata_taxonomy <- BIEN_taxonomy_species(species = "Artemisia tridentata")
Artemisia_tridentata_taxonomy[1:10,]
```
```{r}
Carpinus_caroliniana_taxonomy <- BIEN_taxonomy_species(species = "Carpinus caroliniana")
Carpinus_caroliniana_taxonomy[1:5,]
```

```{r}
Juniperus_virginiana_taxonomy <- BIEN_taxonomy_species(species = "Juniperus virginiana")
Juniperus_virginiana_taxonomy[1:nrow(Juniperus_virginiana_taxonomy),]
```
```{r}
Quercus_laevis_taxonomy <- BIEN_taxonomy_species(species = "Quercus laevis")
Quercus_laevis_taxonomy[1:nrow(Quercus_laevis_taxonomy),]
```

