---
title: "Add Plant Functional Type (PFT) labels"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Plant functional types (PFTs) of interest: coniferous forest, deciduous forest, woody shrub, grass/herb, and bare ground.

## Manually add PFTs to spectral library

```{r load_cleaned_spectra, message=FALSE}
# read cleaned spectral library file 
cleaned_spectra <- read_csv(here::here("analysis", "data", "derived_data", "cleaned_spectra.csv"))
```


```{r}
# list unique taxon labels, add an empty columns to fill with plant functional type (PFT) and common name labels.
# PFTs of interest: coniferous_forest, deciduous_forest, woody_shrub, grass_herb, and bare_ground.
taxon_pft_lookup <- data.frame(genusSpecies = unique(cleaned_spectra$genusSpecies),
                               commonName = NA,
                               pft = NA)

for(i in 1:nrow(taxon_pft_lookup)){
  # print row and current genusSpecies label
  print(i)
  genusSpecies <- taxon_pft_lookup$genusSpecies[i]
  print(genusSpecies)
  
  if(genusSpecies == "Pseudotsuga menziesii"){
    taxon_pft_lookup$pft[i] = "coniferous_forest"
    taxon_pft_lookup$commonName[i] = "Douglas-fir"} 
  
  else if(genusSpecies == "Alnus rubra"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"  # deciduous broadleaf tree 
    taxon_pft_lookup$commonName[i] = "red alder"}
  
  else if(genusSpecies == "Vaccinium parvifolium"){ 
    taxon_pft_lookup$pft[i] = "woody_shrub"       # deciduous Shrub
    taxon_pft_lookup$commonName[i] = "red huckleberry"}
  
  else if(genusSpecies == "Mahonia nervosa"){
    taxon_pft_lookup$pft[i] = "woody_shrub"      # evergreen shrub
    taxon_pft_lookup$commonName[i] = "Cascade barberry"}
  
  else if(genusSpecies == "Corylus cornuta"){
    taxon_pft_lookup$pft[i] = "woody_shrub"      # deciduous shrubby hazel
    taxon_pft_lookup$commonName[i] = "beaked hazelnut"}

  else if(genusSpecies == "Gaultheria shallon"){
    taxon_pft_lookup$pft[i] = "woody_shrub"     # broadleaved evergreen shrub
    taxon_pft_lookup$commonName[i] = "salal"}
  
  else if(genusSpecies == "Acer circinatum"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
                  # or woody shrub? USDA Plants shows both Shrub and Tree growth form
    taxon_pft_lookup$commonName[i] = "vine maple"}
  
  else if(genusSpecies == "Frangula purshiana"){
    taxon_pft_lookup$pft[i] = "woody_shrub" 
    taxon_pft_lookup$commonName[i] = "Cascara buckthorn"}
  
  else if(genusSpecies == "Acer rubrum"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$commonName[i] = "red maple"}

  else if(genusSpecies == "Betula alleghaniensis"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$commonName[i] = "yellow birch"}
  
  else if(genusSpecies == "Populus tremuloides"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$commonName[i] = "quaking aspen"}
  
  else if(genusSpecies == "Tsuga canadensis"){
    taxon_pft_lookup$pft[i] = "coniferous_forest"
    taxon_pft_lookup$commonName[i] = "eastern hemlock"}
  
  else if(genusSpecies == "Acer pensylvanicum"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$commonName[i] = "striped maple"}
  
  else if(genusSpecies == "Picea sp."){
    taxon_pft_lookup$pft[i] = "coniferous_forest"
    taxon_pft_lookup$commonName[i] = "spruce"}
  
  else if(genusSpecies == "Fraxinus americana"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$commonName[i] = "white ash"}
  
  else if(genusSpecies == "Betula papyrifera"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$commonName[i] = "paper birch"}
  
  else if(genusSpecies == "Acer saccharum"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$commonName[i] = "sugar maple"}
  
  else if(genusSpecies == "Fagus grandifolia"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$commonName[i] = "American beech"}
  
  else if(genusSpecies == "Rhamnus davurica"){
    taxon_pft_lookup$pft[i] = "woody_shrub"
    taxon_pft_lookup$commonName[i] = "Dahurian buckthorn"}
  
  else if(genusSpecies == "Lonicera maackii"){
    taxon_pft_lookup$pft[i] = "woody_shrub"
    taxon_pft_lookup$commonName[i] = "Amur honeysuckle"}
  
  else if(genusSpecies == "Ailanthus altissima"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$commonName[i] = "tree of heaven"}
  
  else if(genusSpecies == "Prunus serotina"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$commonName[i] = "black cherrt"}
  
  else if(genusSpecies == "Celtis occidentalis"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$commonName[i] = "common hackberry"}
  
  else if(genusSpecies == "Quercus alba"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$commonName[i] = "white oak"}
  
  else if(genusSpecies == "Juglans nigra"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$commonName[i] = "black walnut"}
  
  else if(genusSpecies == "Robinia pseudoacacia"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$commonName[i] = "black locust"}
  
  else if(genusSpecies == "Pinus strobus"){
    taxon_pft_lookup$pft[i] = "coniferous_forest"
    taxon_pft_lookup$commonName[i] = "eastern white pine"}
  
  else if(genusSpecies == "Carya glabra"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$commonName[i] = "pignut hickory"}
  
  else if(genusSpecies == "Tetradium daniellii"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$commonName[i] = "bee-bee tree"}
  
  else if(genusSpecies == "Juniperus virginiana"){
    taxon_pft_lookup$pft[i] = "coniferous_forest"
    taxon_pft_lookup$commonName[i] = "red cedar"}
  
  else if(genusSpecies == "Picea mariana"){
    taxon_pft_lookup$pft[i] = "coniferous_forest"
    taxon_pft_lookup$commonName[i] = "black spruce"}
  
  else if(genusSpecies == ""){
    taxon_pft_lookup$pft[i] = ""
    taxon_pft_lookup$commonName[i] = ""}
  
  else if(genusSpecies == ""){
    taxon_pft_lookup$pft[i] = ""
    taxon_pft_lookup$commonName[i] = ""}
  
  else if(genusSpecies == ""){
    taxon_pft_lookup$pft[i] = ""
    taxon_pft_lookup$commonName[i] = ""}
  
  else if(genusSpecies == ""){
    taxon_pft_lookup$pft[i] = ""
    taxon_pft_lookup$commonName[i] = ""}
  
  else if(genusSpecies == ""){
    taxon_pft_lookup$pft[i] = ""
    taxon_pft_lookup$commonName[i] = ""}
  
  else print("PFT unknown")
  
  # print corresponding PFT
  print(taxon_pft_lookup$pft[i])
}

taxon_pft_lookup$pft(taxon_pft_lookup$genusSpecies == "Pseudotsuga menziesii") <- "coniferous_forest"
```

## R packages: attempting to automate PFT labeling

### USDA Plants Database

USDA Plants Database R API tutorial: https://recology.info/2016/10/usda-plants-database-r/ 

The USDA maintains a database of plant information, some of it trait data, some of it life history. Check it out at http://plants.usda.gov/java/

```{r install_load_packages, message=FALSE}
#install.packages("request")
#library(devtools)
#devtools::install_github("sckott/request")
library(request)
library(tibble)
library(tidyverse)
library(here)
```

```{r}
# Set the base url
root <- api("https://plantsdb.xyz")

root %>% api_path(heartbeat)
```

^^ My results here do not match those that show up on the tutorial page above. Maybe the API is down or the URL has changed? 

Based on the GitHub issue discussion (https://github.com/USDA/USDA-APIs/issues/7), someone uploaded USDA plants data here from 2016: https://uofi.app.box.com/v/usdaplants

```{r, eval=FALSE, include=FALSE}
# Specify URL where file is stored 
url <- "https://uofi.app.box.com/v/usdaplants/folder/6688524249/usda_plants_20160713.csv"
destFile <- file.path(here::here(), "analysis","data", "usda_plants_20160713.csv")
# Apply download.file function in R
download.file(url, destFile)
```

```{r, include = FALSE, eval=FALSE}
#usda_plants <- read_csv(file.path(here::here(), "analysis", "data", "usda_plants_20160713.csv"))

# Error in nchar(x) : invalid multibyte string, element 2
# ^ Issue with some of the column names in the .csv? 
```



### BIEN 

RBIEN tutorial: https://cmerow.github.io/RDataScience/3_3_RBIEN_tutorial.html

```{r install_load_packages_bien, message=FALSE, warning=FALSE}
library(BIEN)
library(ape) # Package for working with phylogenies in R
library(maps) # Useful for making quick maps of occurrences
library(sp) # A package for spatial data
```

BIEN traits and taxonomic data

"Let’s say we’re interested in the genus [Abies], and we’d like to get an idea of how many species there are in this genus and what higher taxa it falls within."

*Abies* genus is fir trees (coniferous trees). The "higher_plant_group" returned by the BIEN taxonomy function is "gymnosperms (conifers)"

```{r}
Abies_taxonomy<-BIEN_taxonomy_genus(genus = "Abies")

#View(Abies_taxonomy)
Abies_taxonomy[1:3,]
```

*Acer rubrum* species is the red maple (deciduous trees). The "higher_plant_group" returned by the BIEN taxonomy is "flowering plants".

```{r}
Acer_rubrum_taxonomy <- BIEN_taxonomy_species(species = "Acer rubrum")
Acer_rubrum_taxonomy[1:3,]
```

### TRY Database

https://www.try-db.org/TryWeb/Home.php

### Taxize

https://www.rdocumentation.org/packages/taxize/versions/0.9.99
